### 2015-11-20 paper.Rmd
update with new plots
* some corrections
* add Jeff to authors
* PDF plots (needs mrds/Distance from github)

---
author:
  - name: David L Miller
    affiliation: Centre for Research into Ecological and Environmental Modelling, University of St Andrews
    address: >
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: dave@ninepointeightone.net
    url: http://converged.yt
title:
  formatted: "Distance Sampling in \\proglang{R}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Distance Sampling in R"
  # For running headers, if needed
  #short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [distance sampling, abundance estimation, line transects, point transects, "\\proglang{R}"]
  plain:     [distance sampling, abundance estimation, line transects, point transects, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{booktabs}
output: rticles::jss_article
bibliography: jstatsoft.bib
---

@@

Looking at an example of the data format given in the last bullet, we can examine the minke whale data:

```{r load-data, echo=FALSE}
# quietly load the data
load("data/minke.RData")
load("data/amakihi.RData")
```
```{r minke-data-head, message=FALSE}
library(Distance)
head(minke)
```
and the amakihi data:
```{r amakihi-data-head}
head(amakihi)
```

* what do we want to say here?

@@

As mentioned above, the detection function describes the relationship between observed distances and probability of detection. The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance and $\boldsymbol{\theta}$ is a vector of parmeters to be estimated. Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over the distances), so we must integrate out distance ($y$) from the detection function:
$$
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
$$
where $\pi(y)$ describes the distribution of objects with respect to the sampler and is $1/w$ for line transects and $\frac{2r}{w}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3].

@@

There is a wide literature on possible formulations for the detection function [@Buckland:1992fa; @Eidous:2005bj; @Becker:2009cj; @Giammarino:2014eg; @Miller:2015hw; @Becker:2015fi]. \code{Distance} includes the most popular of these models, and includes an extendable class system to add new detection functions while avoiding code duplication (see "Extensions").

Here we'll detail the most popular detection function approach: "key function plus adjustments" (K+A).

@@

Key function plus adjustment terms (or adjustment series) models are formulated by taking a "key" function and optionally adding "adjustments" to it to improve the fit. Mathematically we formulate this as:
$$
g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
$$
where $k$ is the key function and $\alpha_O$ is sum series of functions (given in Table \ref{tab:keyadj}), described as an *adjustment of order $O$*. Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e. $\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})$).

Some models for the key are as follows:
$$
k(y) = \left\{
\begin{array}{l l}
  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal,} \\
  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate,} \\
  1/w & \quad \text{uniform.}
\end{array} \right.
$$
Possible modelling options for key and adjustments are given in Table \ref{tab:keyadj} and illustrated in Figure \ref{fig:hnhr} and \ref{fig:keyadj}. We select the number of adjustment terms ($K$) by AIC (further details in "Model checking and model selection").

@@

We do not always use adjustments (except in the case of the uniform key), in which case we refer to "key only" models; see also "Covariates" and "Model checking and model selection" below.

@@

Figure \ref{fig:minkeamakihi} (left panel) shows the result of calling \code{plot} on the resulting model object. We can also call \code{summary} on the model object to get summary information about the fitted model (though we postpone this until the next section).

@@

By default \code{ds} assumes that the data given to it is line transect, but we can switch to points using \code{transect="point"}. Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate (and truncating at 82.5m, given in @Marques:2007ey) can be specified by:
```{r amakihi-hr-obs}
amakihi_hr_obs <- ds(amakihi, truncation=82.5, transect="point",
                     key="hr", formula=~obs)
```
As with the minke whale model, we can plot the resulting detection function (Figure \ref{fig:minkeamakihi}, centre and right panels). *Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate and separately for each level set (for continuous covariates, the 25\%, 50\% and 75\% quantiles are displayed).*

**It would be nice if I could get "nice" plotting working for `Distance` as they are much easier to interpret (IMO)**

*When multiple covariates are used in the model, a series of plots are produced. We can see this in the bottom row of Figure \ref{fig:minkeamakihi}, where we plot the results of fitting following model:*
```{r amakihi-hr-obs-mas, warning=FALSE}
amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
                         key="hr", formula=~obs+mas)
```

```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=3, fig.cap="Plot of fitted detection functions overlayed on the histograms of observed distances. Left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates.\\label{fig:minkeamakihi}"}
par(mfrow=c(1,3))
plot(minke_hn)
plot(amakihi_hr_obs)
plot(amakihi_hr_obs_mas)
```

@@

As with models fitted using \code{lm} or \code{glm} in \proglang{R} we can use \code{summary} to give us some useful information about our fitted model. For example for our hazard-rate model for the amakihi, with observer as a covariate:
```{r amakihi-summary}
summary(amakihi_hr_obs)
```
This summary information includes various details of the data and model specification, as well as the values of the coefficents ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and it's uncertainty. The final line gives an estimate of abundance for the area covered by the survey (again, this will be covered in detail in the next section).

@@

### Post-stratification

@@

>>>>

---
author:
  - name: David L Miller
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: dave@ninepointeightone.net
    url: http://converged.yt
  - name: Jeffrey L Laake
    affiliation: National Marine Mammal Laboratory
    address: >
      Alaska Fisheries Science Center
      7600 Sand Point Way N.E., Seattle, WA 98115, USA
    email: Jeff.Laake@noaa.gov
title:
  formatted: "Distance Sampling in \\proglang{R}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Distance Sampling in R"
  # For running headers, if needed
  #short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [distance sampling, abundance estimation, line transects, point transects, "\\proglang{R}"]
  plain:     [distance sampling, abundance estimation, line transects, point transects, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{booktabs}
output: rticles::jss_article
bibliography: jstatsoft.bib
---

@@

The minke whale data follows the "flatfile" format given in the last bullet point:
```{r load-data, echo=FALSE}
# quietly load the data
load("data/minke.RData")
load("data/amakihi.RData")
```
```{r minke-data-head, message=FALSE}
library(Distance)
head(minke)
```
Where as the amakihi data lacks effort and stratum data:
```{r amakihi-data-head}
head(amakihi)
```

We'll explore what the effects of including effort data are during analysis below.

@@

As mentioned above, the detection function describes the relationship between observed distances and probability of detection. The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (either from a line or point) and $\boldsymbol{\theta}$ is a vector of parmeters to be estimated. Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over the distances), so we must integrate out distance ($y$) from the detection function:
$$
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
$$
where $\pi(y)$ describes the distribution of objects with respect to the sampler and $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3] (letting $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point).

@@

There is a wide literature on possible formulations for the detection function [@Buckland:1992fa; @Eidous:2005bj; @Becker:2009cj; @Giammarino:2014eg; @Miller:2015hw; @Becker:2015fi]. \code{Distance} includes the most popular of these models. Here we'll detail the most popular detection function approach: "key function plus adjustments" (K+A).

@@

Key function plus adjustment terms (or adjustment series) models are formulated by taking a "key" function and optionally adding "adjustments" to it to improve the fit [@Buckland:1992fa]. Mathematically we formulate this as:
$$
g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
$$
where $k$ is the key function and $\alpha_O$ is sum series of functions (given in Table \ref{tab:keyadj}), described as an *adjustment of order $O$*. Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e. $\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})$).

Available models for the key are as follows:
$$
k(y) = \left\{
\begin{array}{l l}
  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal,} \\
  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate,} \\
  1/w & \quad \text{uniform.}
\end{array} \right.
$$
Possible modelling options for key and adjustments are given in Table \ref{tab:keyadj} and illustrated in Figure \ref{fig:hnhr} and \ref{fig:keyadj}. We select the number of adjustment terms ($K$) by AIC (further details in "Model checking and model selection").

@@

We do not always include adjustments (except in the case of the uniform key), in which case we refer to "key only" models (see the next section and "Model checking and model selection").

@@

Figure \ref{fig:minkeamakihi} (top left plot) shows the result of calling \code{plot} on the resulting model object. We can also call \code{summary} on the model object to get summary information about the fitted model (though we postpone this until the next section).

@@

By default \code{ds} assumes that the data given to it is line transect, but we can switch to points using \code{transect="point"}. Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate can be specified by [truncating at 82.5m, given in @Marques:2007ey]:
```{r amakihi-hr-obs}
amakihi_hr_obs <- ds(amakihi, truncation=82.5, transect="point",
                     key="hr", formula=~obs)
```
As with the minke whale model, we can plot the resulting detection function (Figure \ref{fig:minkeamakihi}, centre column). Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate. Points on the plot indicate probability of detection  for each observation. For the `amakihi_hr_obs` model we see fairly clear levels of the observer covariate in the points. Looking at the left panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear when adding minutes after sunrise as a covariate into the model:
```{r amakihi-hr-obs-mas, warning=FALSE}
amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
                         key="hr", formula=~obs+mas)
```

The bottom row of Figure \ref{fig:minkeamakihi} shows the probability density functions for the amakihi models. These often are easier to interpret than the detection functions, as for point data we must rescale the histogram when plotting the detection function to take into account the geometry of the point sampler.

```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=6, fig.cap="Top row plots of fitted detection functions overlayed on the histograms of observed distances; left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates (for amakihi histograms are rescale to take into account sampler geometry). Bottom row shows plots of the probability density function for the amakihi models (as top row). For covariate models (centre and right columns) points indicate probability of detection for a given observation (given that observations covariate values) and lines indicate the average detection function.\\label{fig:minkeamakihi}"}
par(mfrow=c(2,3))
plot(minke_hn, showpoints=FALSE)
plot(amakihi_hr_obs)
plot(amakihi_hr_obs_mas)
plot(1,1, type="n", axes=FALSE, xlab="", ylab="")
plot(amakihi_hr_obs, pdf=TRUE)
plot(amakihi_hr_obs_mas, pdf=TRUE)
```

@@

As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give us some useful information about our fitted model. For example for our hazard-rate model for the amakihi, with observer as a covariate:
```{r amakihi-summary}
summary(amakihi_hr_obs)
```
This summary information includes various details of the data and model specification, as well as the values of the coefficents ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and it's uncertainty. The final line gives an estimate of abundance for the area covered by the survey (again, this will be covered in detail in the next section).

@@

### Stratification

@@

We combine these two sources of variance by noting that squared coefficients of variation (approximately) add [@Goodman:1960dd].
